A comprehensive framework to capture the arcana of neuroimaging analysis

### 2019-03-18 manuscript/manuscript.tex
touched up methods and started writing FAIR paragraph in discussion

All MultiStudy classes must inherit from the Multi\emph{Study} base
class and be created by the Multi\emph{StudyMetaClass} metaclass or
subclasses thereof. As in the case of subclassing the standard Study
class, additional data and parameter specifications can be added to the
class via \emph{add\_data\_specs} and \emph{add\_param\_specs}
respectively for additional analysis not included in the sub-studies.

Sub-studies are aggregated in the \emph{sub-study} \emph{specification}
of a MultiStudy class via a list of \emph{SubStudySpec} objects
in the \emph{add\_sub\_study\_specs} class attribute in the manner of
data and parameter specifications. A SubStudySpec consists of a name, a
Study class, and a \emph{name map} dictionary\emph{.} The name map
dictionary maps data and parameter specification names from the
sub-study namespace to the namespace of the MultiStudy class, i.e.
the dictionary keys refer to entries in the sub-study specification and
the dictionary values refer to entries in the multi-study specification.

Entries in the specifications of sub-study classes that are not
referenced in the sub study's name map are implicitly mapped to the
MultiStudy namespace by the MultiStudyMetaClass during construction of the
MultiStudy class using the name of the sub-study as a prefix.
If the implicitly mapped specification is derived, then its associated pipeline
constructor method is also mapped into the MultiStudy namespace with a prefix.
For example, \emph{derived1} in \emph{sub\_study2} would be mapped to
\emph{sub\_study2\_derived1} along with the method
\emph{sub\_study2\_pipeline1}.

@@

To apply the analysis blueprint specified in a Study class to a specific
dataset, an instance of the Study class is created with details of where
the data is stored (\emph{Repository} module) the computing
resources available to process it (\emph{Processor} module) and the software
installed in the environment it will be processed in (\emph{Environment} module).
A Study object controls the construction and execution of analysis workflows,
and the flow of data to and from the repository (Figure \ref{fig:full_uml}).

@@

Each Study instance is assigned a name, which is used to differentiate its
results from alternative analyses on the same dataset (e.g. with different
parameterisations). Parameters are set on initialisation of the Study object
along with the range of subject and visit IDs to be included in the analysis
(if they are not provided then all IDs found in the repository are included).
The remaining arguments passed to the Study object initialisation are the
Repository, Processor and Environment modules to use and a list of
\emph{Input} objects to select input data from the repository and match
it with the Study's data specification.
(Figure \ref{fig:study_application}).

@@

In its most basic form, a ``basic'' repository, is just a file system
directory containing the data to be processed for a single subject.
For multi-subject studies, the root directory should contain separate
subdirectories for each subject in the study, with subject IDs taken
from the subdirectory names. If data was acquired for each subject
over multiple visits, than an additional layer of nested subdirectories
is included in the subject subdirectories, with the visit IDs taken from
the subdirectory names. Note that the basic
repository is similar to the BIDS format, however, there are no naming
conventions in the basic repository, which enables prototyping and
testing of analyses on loosely structured data.

Derivatives are stored by their specification name in a study-specific
namespaces to avoid clashes with separate analyses. In basic
repositories this namespace is a subdirectory named after the study
nested within the lowest layer of the data tree. In BIDS repositories, the namespace is a subdirectory of the
\emph{derivatives} directory, again named after the study. In XNAT
repositories, derivatives for each session are stored in separate
\emph{MrSession} objects alongside the primary session underneath its
\emph{Subject}, and are named
\emph{\textless{}primary-session-name\textgreater{}\_\textless{}study-name\textgreater{}}
(Table \ref{tbl:repo-structure}).

Derived filesets are stored with the format specified in the study's
data specification. In plain-directory and BIDS repositories, fields are
stored in a single JSON file named `\_\_fields\_\_.json' in each derived
session, and on XNAT they are stored in custom fields of the derived
session. Provenance is stored in a `\_\_prov\_\_' sub-directory (dataset
on XNAT) of the derivatives directory (MrSession on XNAT) in separate
JSON files for each pipeline named after the pipeline (Table 1).

Summary data (i.e. with \emph{per\_subject}, \emph{per\_visit}, and
\emph{per\_study} frequencies\emph{)} are stored in specially named
subjects and visits (e.g. `group'), the names for which are specified when
the repository is initialised. For example, given in plain-directory
repository using all as the summary name for both subjects and visits,
per\_subject data for `subj1' would be stored at
\emph{\textless{}root\textgreater{}/subj1/}group, per\_visit data for
`visit1' in \emph{\textless{}root\textgreater{}/}group\emph{/visit1},
and per\_study data in \\ \emph{\textless{}root\textgreater{}/}group/group
(Table \ref{tbl:repo-structure}).

@@

While derivatives generated by a Study object are named in accordance
with the data specification of the Study class, arbitrary naming
conventions can be used for input datasets and fields. A selection
stage matches input data to entries in the data specification of
the Study class. The criteria for these selections are passed to the
the \emph{inputs} argument of the Study object at instantiation in a
dictionary mapping data specification names to \emph{FilesetInput} and
\emph{FieldInput} objects, and are required
to match exactly one fileset or field in every session included in the study.

Matching is typically performed on file
names (dataset labels for XNAT repositories) and field names. If the
names are inconsistent across the study then regular expressions can be
used instead of exact matches with the \emph{pattern} keyword argument.
Additional criteria can be used to distinguish cases where multiple
filesets in the session match the pattern provided, such as DICOM
header values, or the order and ID of the dataset.

@@

If the data to select has been derived from an alternative Arcana Study
instance, then the name of the alternative study can be passed to the
\emph{from\_study} keyword argument provided to the selector. There are
no restrictions on selecting any data, derived or otherwise, to match
input or derived specifications. For example, it is possible to
deliberately skip analysis steps by selecting an output of an early step
in a previous run as a match for a later derived spec although this is
not recommended as standard practice.

Specific files and fields that are not stored within a repository can be
passed as inputs in \emph{FilesetCollection} and \emph{FieldCollection}
objects. Collection objects reference an entry in the data specification
and contain a single Fileset or Field for every session (or
subject/visit/study depending on the frequency of the corresponding data
specifcation). Collection objects can be used to pass reference atlases and
templates as inputs to analyses. They can also be set as the default
for a data specification via the \emph{default} keyword
argument. However, for the sake of portability, default inputs should be
restricted to data in publically accessible repositories or those included in
standard software packages (e.g. FSL).

When using BIDS repositories, the selection stage is typically already
included in the data specification (see \emph{Data and parameter
specifications)} so inputs do not need to be provided to the
initialisation of the Study. However, BidsInput and
BidsAssocInput objects can be provided to override the default
selections if required.

@@

Provenance is stored for each pipeline run alongside the generated
derivatives and consists of parameter values used by the pipeline,
software versions used by the pipeline, interface parameters,
a graph representation of the underlying Nipype workflow, checksums of inputs,
version of Arcana used, version of Nipype used, and any subject and visit IDs
that were joined over in the workflow.

For subsequent analyses, changes with respect to any of the stored provenance values will
be flagged as a mismatch, with the exception of interface package versions
(e.g. Nipype or Banana versions). How mismatches are handled depends on the \emph{reprocess} flag passed to the
Processor. If \emph{reprocess} is true then the sessions with mismatching provenance
will be reprocessed, otherwise if \emph{reprocess} is false (the default) an exception will be raised.
Changes with respect to any element in the provenance JSON document can be ignored by providing
a list of JSONPath queries to the \emph{prov\_ignore} keyword argument.

@@

The Arcana framework is distributed as a publicly available software
package via GitHub (\url{github.com/MonashBI/arcana}) and the
Python Package Index (PyPI) (\url{pypi.org/project/arcana/}).
Study classes for T1, T2* and diffusion weighted MRI data have been
implemented as part of the \emph{Biomedical imging ANAlysis iN Arcana (Banana)} package
(\url{github.com/MonashBI/banana}; \url{pypi.org./project/banana}).
All three classes, \emph{T1Study}, \emph{T2starStudy} and
\emph{DwiStudy}, inherit generic image analysis methods, such as
registration and brain extraction, from the base class \emph{MriStudy}.

@@

>>>>

All MultiStudy classes must inherit from the Multi\emph{Study} base
class and be created by the \emph{MultiStudyMetaClass} metaclass or
subclasses thereof. As in the case of subclassing the standard Study
class, additional data and parameter specifications can be added to the
class via \emph{add\_data\_specs} and \emph{add\_param\_specs}
respectively for additional analysis not included in the sub-studies.

Sub-studies are aggregated in the \emph{sub-study} \emph{specification}
of a MultiStudy class via a list of \emph{SubStudySpec} objects
in the \emph{add\_substudy\_specs} class attribute in the manner of
data and parameter specifications. A SubStudySpec consists of a name, a
Study class, and a \emph{name map} dictionary\emph{.} The name map
dictionary maps data and parameter specification names from the
sub-study namespace to the namespace of the MultiStudy class, i.e.
the dictionary keys refer to entries in the sub-study specification and
the dictionary values refer to entries in the multi-study specification.

Entries in the specifications of sub-study classes that are not
referenced in the sub study's name map are implicitly mapped to the
MultiStudy namespace by the MultiStudyMetaClass during construction of the
MultiStudy class using the name of the sub-study as a prefix.
If the implicitly mapped specification is derived, then its associated pipeline
constructor method is also mapped into the MultiStudy namespace with a prefix.
For example, \emph{derived1} in \emph{substudy2} would be mapped to
\emph{substudy2\_derived1} along with the method
\emph{substudy2\_pipeline1}.

@@

To apply the analysis blueprint specified in a Study class to a specific
dataset, an instance of the Study class is created with details of where
the data is stored (\emph{Repository} module and \emph{Input} selectors) the computing
resources available to process it (\emph{Processor} module) and the software
installed in the environment it will be processed in (\emph{Environment} module).
A Study object controls the construction and execution of analysis workflows,
and the flow of data to and from the repository (Figure \ref{fig:full_uml}).

@@

Each Study instance is assigned a name, which is used to differentiate its
results from alternative analyses on the same dataset (e.g. with different
parameterisations). Parameters are set on initialisation of the Study object
along with the range of subject and visit IDs to be included in the analysis
If subject and visit IDs are not explicitly provided then all IDs found in the repository are included.
The remaining arguments passed to the Study object at initialisation are the
Repository, Processor and Environment modules and a list of
\emph{Input} objects to match data in the repository with inputs in the
data specification (Figure \ref{fig:study_application}).

@@

In its most basic form, a ``basic'' repository, is just a file system
directory containing the data to be processed for a single subject.
For multi-subject studies, the root directory should contain separate
subdirectories for each subject in the study, with subject IDs taken
from the subdirectory names. If data was acquired for each subject
over multiple visits, than an additional layer of nested subdirectories
is included in the subject subdirectories, with the visit IDs taken from
the subdirectory names. The basic
repository is similar to the BIDS format in this form, however, there are no naming
conventions in the basic repository, which enables prototyping and
testing of analyses on loosely structured data.

Derivatives are stored by their specification name in a study-specific
namespaces to avoid clashes between separate analyses. In basic
repositories this namespace is a subdirectory named after the study
nested within the lowest layer of the data tree. In BIDS repositories, the namespace is a subdirectory of the
\emph{derivatives} directory, again named after the study. In XNAT
repositories, derivatives for each session are stored in separate
\emph{MrSession} objects alongside the primary session underneath its
\emph{Subject}, and are named
\emph{\textless{}primary-session-name\textgreater{}\_\textless{}study-name\textgreater{}}
(Table \ref{tbl:repo-structure}).

Derived filesets are stored with the format specified in the study's
data specification. In basic and BIDS repositories, fields are
stored in a single JSON file named `\_\_fields\_\_.json' in each derived
session, and on XNAT they are stored in custom variables in the derived
session. Provenance is stored in a `\_\_prov\_\_' sub-directory (dataset
on XNAT) of the derivatives directory (MrSession on XNAT) in separate
JSON files for each pipeline (Table 1).

Data from group analyses (i.e. with \emph{per\_subject}, \emph{per\_visit}, and
\emph{per\_study} frequencies\emph{)} are stored in specially named
subjects and visits (e.g. `group'), the names for which are specified when
the repository is initialised. For example, given a basic
repository using `group' as the summary name for both subjects and visits,
per\_subject data for `subj1' would be stored at
\emph{\textless{}root\textgreater{}/subj1/}group, per\_visit data for
`visit1' in \emph{\textless{}root\textgreater{}/}group\emph{/visit1},
and per\_study data in \\ \emph{\textless{}root\textgreater{}/}group/group
(Table \ref{tbl:repo-structure}).

@@

While derivatives generated by a Study object are named in accordance
with the data specification of the Study class, input file sets and fields
do not need to be conventionally named when using basic or XNAT
repositories. In these cases, a selection stage matches input data to
entries in the data specification of the Study class. The criteria for
these selections are passed to the
the \emph{inputs} argument of the Study object at instantiation in a
dictionary mapping data specification names to \emph{FilesetInput} and
\emph{FieldInput} objects.

FilesetInput and FieldInput objects are required
to match exactly one fileset or field in every session included in the study.
Matching is typically performed on file
names (dataset labels for XNAT repositories) and field names. If the
names are inconsistent across the study then regular expressions can be
used instead of exact matches with the \emph{pattern} keyword argument.
Additional criteria can be used to distinguish cases where multiple
filesets in the session match the pattern provided, such as DICOM
header values, or the order and ID of the dataset.

@@

If the data to select has been derived from an alternative Study
instance, then the name of the alternative study can be passed to the
\emph{from\_study} keyword argument provided to the Input object. There are
no restrictions on selecting any data, derived or otherwise, to match
input or derived specifications. For example, it is possible to
deliberately skip analysis steps by selecting an upstream derivative
from a previously run workflow as a match for a downstream derivative, although this is
not recommended as standard practice.

Specific file sets and fields can be passed as inputs to a Study in
\emph{FilesetCollection} and \emph{FieldCollection}
objects. Collection objects reference an entry in the data specification
and contain a single Fileset or Field for every session (or
subject/visit/study depending on the frequency of the corresponding data
specifcation). Collection objects can be used to pass reference atlases and
templates as inputs to analyses. They can also be set as the default
for a data specification via the \emph{default} keyword
argument. However, for the sake of portability, default inputs should be
restricted to data in publically accessible repositories or those included in
standard software packages (e.g. FSL).

When using BIDS repositories, the input selection stage is typically
included in the data specification (see \emph{Data and parameter
specifications)} so inputs do not need to be provided to the
initialisation of the Study. However, BidsInput and
BidsAssocInput objects can be provided to override the default
selections if required.

@@

Provenance is stored for each pipeline run alongside the generated
derivatives and consists of interface parameter values,
software versions of external requirements, Arcana and Nipype, 
a graph representation of the underlying Nipype workflow, checksums of inputs
and outputs, and any subject and visit IDs that were joined over in the workflow.

For subsequent analyses, changes with respect to any of the stored provenance values will
be flagged as a mismatch by default, with the exception of interface package versions
(e.g. Nipype or Banana versions). How mismatches are handled depends on the \emph{reprocess} flag passed to the
Processor. If \emph{reprocess} is true then the sessions with mismatching provenance
will be reprocessed, otherwise if \emph{reprocess} is false (the default) an exception will be raised.
Changes with respect to any element in the provenance JSON document can be ignored by providing
a list of JSONPath queries to the \emph{prov\_ignore} keyword argument.

@@

The Arcana framework is distributed as a publicly available software
package via GitHub (\url{github.com/MonashBI/arcana}) and the
Python Package Index (PyPI) (\url{pypi.org/project/arcana/}).
Study classes for T1, T2* and diffusion weighted MRI data have been
implemented as part of the \emph{Biomedical imging ANAlysis iN Arcana (Banana)} package
(\url{github.com/MonashBI/banana}; \url{pypi.org/project/banana}).
All three classes, \emph{T1Study}, \emph{T2starStudy} and
\emph{DwiStudy}, inherit generic image analysis methods, such as
registration and brain extraction, from the base class \emph{MriStudy}.

@@

The framework that Arcana provides should facilitate the development of workflows
that are consistent with the F.A.I.R. guiding principles \citep{wilkinson_fair_2016}.
\begin{itemize}
\item Findable all in one code-base
\item Accessible open source and implemented in different systems
\item Interoperable with BIDS derivatives
\item Interoperable with different analysis software
\item Reusable in different contexts
\item Reusable with detailed provenance
\end{itemize}
